### 2017-06-29 motivation.tex
Edits.

While the aforementioned reasoning is simple to perform manually, the
sequential verification of database programs, such as the one in
Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
presence of loops and collections. Thus verifying serializable
transactions is at least as difficult as reasoning about sequential
imperative programs. Nonetheless, program logics and automation
techniques developed to reason about imperative programs can be used
to reason about sequential transactions, perhaps with reasonable
success.

concurrent withdraw transactions on an account balance (\C{B}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficient
condition to preserve invariants. But, is {\sc ser} necessary?

As an alternative, consider the execution of this transaction under a
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} of uncommitted data. In
the current example, {\sc rc} isolation admits the two executions
shown in Fig.~\ref{fig:rc-ex} on a strongly consistent ({\sc sc})
store, such as a conventional RDBMS.

@@

The figure depicts an execution as a series of read, write, and commit
operations. In the execution on left, transaction \C{Wd1} (green)
reads the current balance (\C{k}) and writes the new balance
(\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes
and commits, writing the new balance (\C{k-a2}). {\sc rc} isolation
prevent \C{Wd2} from witnessing the uncommitted writes of transaction
\C{Wd1}.  Subsequently committing \C{Wd1} leads to the loss of
\C{Wd2}'s updates (the so-called \emph{lost update
anomaly}~\cite{berenson}), resulting in an incorrect balance of
\C{k-a1}. The execution on right describes a similar scenario with
\C{Wd1} and \C{Wd2} exchanging their roles.  Clearly, {\sc rc} is an
excessively weak isolation level for this program because it loses the
updates of one transaction, resulting in the violation of the
post-condition.  A stronger isolation level that prevents lost updates
is required. \iso{Snapshot Isolation} ({\sc si})~\cite{berenson} fits
this requirement; {\sc si} effectively serializes transactions that
update a shared data object by aborting and re-executing a transaction
if write-write conflicts are detected during its commit.\footnote{{\sc
si} however does not serialize transactions in the absence of
write-write conflicts.} Since {\sc si}, unlike {\sc ser}, does not
need expensive mechanisms such as lock-based concurrency control or
runtime monitoring, it is also more efficient, making it appropriate
for both withdraw transactions.

>>>>

While the aforementioned reasoning is simple to perform manually, the
sequential verification of database programs, such as the one in
Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
presence of loops and collections. Thus verification of serializable
transactions is comparable to the verification of sequential
(imperative) programs in terms of difficulty. It is therefore not
unreasonable to expect the reasoning machinery (i.e., program logics
and inference techniques) developed for imperative programs to be
equally effective on serializable transactions.

Unfortunately, transactions are rarely serializable in practice, and
presence of weak isolation significantly complicates the reasoning
about database programs. Weak isolation allows a transaction to
witness the effects of concurrent transactions during its
execution\footnote{Weak isolation doesn't violate atomicity as long as
the witnessed effects are those of the committed transactions}.
Various levels of weak isolation differ in what effects each allows
the transaction to witness, and when. Admitting interference from
concurrent transactions may violate the invariants that would
otherwise hold under a fully isolated execution.

Let us reconsider the \C{new\_order} transaction, which has been
previously shown to preserve TPC-C's consistency condition under a
serializable execution. However, serializability is rarely the default
isolation level on databases. For instance, the default on Postgres is
Read Committed (RC), which is considerably weaker than
serializability. An RC transaction is isolated from \emph{dirty
writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$) trying to place new orders for
different customers under the same district (\C{d\_id}), executions
shown in Fig.~\ref{fig:new_order_execs} are possible on Postgres.

@@

The figure depicts an execution as a series of read, write, and commit
operations. In the execution on left, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases, including Postgres. Post $T_2$'s
commit, $T_1$ resumes execution and adds new \C{Order} and
\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
end of the execution, \C{Order\_line} records inserted by $T_1$ and
$T_2$ all bear the same order id. There are also two \C{Order} records
with the same district id (\C{d\_id}) and order id, none of whose
\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
inserted with that order id. This clearly violates TPC-C's consistency
condition.

Note that, in the above example, there is no Write-Write conflict, or
the \emph{lost update} anomaly that often characterizes concurrency
bugs, such as illegal concurrent withdraws on a bank
account~\cite{poloniexbug}. While $T_1$ and $T_2$ both increment the
\C{d\_next\_o\_id} field of the district record, they do so atomically
(Line 4 of Fig.~\ref{fig:new_order_code}), allowing both updates to be
present in the final state. Likewise, four other well-known anomalies
that characterize RC isolation~\cite{berenson}, namely, \emph{fuzzy
reads}, \emph{phantom reads}, \emph{read skew}, and \emph{write skew},
are also not exhibited by the example. Thus, any program analysis,
static or dynamic, that aims to determine appropriate isolation by
checking for possible manifestations of known anomalies fails to
identify grounds for promoting the isolation level of \C{new\_order}.
Yet, if we take the semantics of the application into account, it is
clear that RC is not an appropriate isolation level for
\C{new\_order}. Clearly, there is a need for a high-level verification
framework that lets us reason about transaction semantics under weak
isolation.

\newpage

(\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes
and commits, writing the new balance (\C{k-a2}). {\sc rc} isolation
prevent \C{Wd2} from witnessing the uncommitted writes of transaction
\C{Wd1}.  Subsequently committing \C{Wd1} leads to the loss of
\C{Wd2}'s updates (the so-called \emph{lost update
anomaly}~\cite{berenson}), resulting in an incorrect balance of
\C{k-a1}. The execution on right describes a similar scenario with
\C{Wd1} and \C{Wd2} exchanging their roles.  Clearly, {\sc rc} is an
excessively weak isolation level for this program because it loses the
updates of one transaction, resulting in the violation of the
post-condition.  A stronger isolation level that prevents lost updates
is required. \iso{Snapshot Isolation} ({\sc si})~\cite{berenson} fits
this requirement; {\sc si} effectively serializes transactions that
update a shared data object by aborting and re-executing a transaction
if write-write conflicts are detected during its commit.\footnote{{\sc
si} however does not serialize transactions in the absence of
write-write conflicts.} Since {\sc si}, unlike {\sc ser}, does not
need expensive mechanisms such as lock-based concurrency control or
runtime monitoring, it is also more efficient, making it appropriate
for both withdraw transactions.
